Abstractions for Multi-Sorted Substitutions

### 2023-09-19 paper/paper.tex
Update.

  In this section, we extend our framework with a single composition
  operator for arbitrary maps.

  The core property of composition is that applying two maps
  \ACode{ϕ₁} and \ACode{ϕ₂} in sequence to a term \ACode{t}, is the same
  as applying their composition \ACode{(ϕ₁ \AField{·ₖ} ϕ₂)} to \ACode{t}, i.e.
  $$
  \ACode{(t \AField{⋯} ϕ₁) \AField{⋯} ϕ₂ \ACon{≡} t \AField{⋯} (ϕ₁ \AField{·ₖ} ϕ₂)}
  $$
  We call this property \ACode{\AField{⋯-fusion}}.
  This property equips our framework with the power to
  reason about the application of multiple maps in terms of
  the application of a single map.
  As such it forms the basis for all lemmas involving multiple maps,
  e.g.\ that applying a weakening and then a singleton
  substitution cancel each other out.

@@

  \subsection{An Examination of Composition}
  \label{sec:composition:example}
  To motivate our abstraction for composition, we first look
  at the four compositions individually:
  \ExampleCode\KExFourComps
  The definitions reveal two interesting properties:
  \begin{enumerate}
  \item
    If we compose two maps \ACode{ϕ₁} and \ACode{ϕ₂}, then the
    resulting map is a renaming, iff both \ACode{ϕ₁} and \ACode{ϕ₂} are
    renamings.
    In other words, 
    if \ACode{ϕ₁} is a \ACode{K₁}-map and \ACode{ϕ₂} is a \ACode{K₂}-map,
    then the result is a \ACode{(K₁ ⊔ K₂)}-map, where \ACode{⊔} refers to the
    lattice for \ACode{\{ \ADef{Kᵣ} , \ADef{Kₛ} \}}
    generated by \ACode{\ADef{Kᵣ} $\leq$ \ADef{Kₛ}}.
  \item
    All four compositions first apply \ACode{ϕ₁} to \ACode{x},
    and then apply \ACode{ϕ₂} to the result.
    If \ACode{ϕ₁} is a renaming, this result is another variable, but
    if \ACode{ϕ₁} is a substitution, this result is a term.
  \end{enumerate}
  With the \ACode{\ACon{Kit}} abstraction, we can easily abstract over
  \ACode{ϕ₂} being a renaming or a substitution:
  \ExampleCode\KExTwoComps

@@

  \LibCode\KComposeKit
  \LibCode\KComposition
  \LibCode\KComposeKitAp
  % \LibCode\KComposeKitApProof
  \LibCode\KDistLiftCompose
  % \LibCode\KDistLiftComposeProof
  \LibCode\KComposeKitNotation
  \LibCode\KComposeTraversal
  \LibCode\KCommLiftWeaken
  % \LibCode\KCommLiftWeakenProof
  \LibCode\KCommLiftWeakenTraverse
  % \LibCode\KCommLiftWeakenTraverseProof
  \LibCode\KComposeKitInstances
  \LibCode\KComposeKitInstancesConcrete
  \LibCode\KWeakenCancelsSingle
  % \LibCode\KWeakenCancelsSingleProof
  \LibCode\KWeakenCancelsSingleTraverse
  % \LibCode\KWeakenCancelsSingleTraverseProof
  \LibCode\KDistLiftSingle
  % \LibCode\KDistLiftSingleProof
  \LibCode\KDistLiftSingleTraverse
  % \LibCode\KDistLiftSingleTraverseProof

  \subsection{System F}
  \AppCode\FFusion
  \AppCode\FFusionProofInteresting
  % \AppCode\FFusionProofRest
  \AppCode\FComposeTraversal

>>>>

  In this section, we extend our framework with an abstraction that
  allows the composition of arbitrary maps.

  The core property of composition, which we call
  \ACode{\AField{⋯-fusion}}, states that applying two maps \ACode{ϕ₁}
  and \ACode{ϕ₂} in sequence to a term \ACode{t}, is the same as
  applying their composition \ACode{(ϕ₁ \AField{·ₖ} ϕ₂)} to \ACode{t},
  i.e.
  $$
  \ACode{(t \AField{⋯} ϕ₁) \AField{⋯} ϕ₂ \ACon{≡} t \AField{⋯} (ϕ₁ \AField{·ₖ} ϕ₂)}
  $$
  This property equips our framework with the power to
  reason about the application of multiple maps in terms of
  the application of a single map.
  As such it forms the basis for all lemmas involving multiple maps,
  e.g.\ that applying a weakening and then a singleton
  substitution cancel each other out.

@@

  \subsection{An Examination of Composition}
  \label{sec:composition:example}
  To motivate our abstraction for composition, we first look
  at the four compositions individually:
  \ExampleCode\KExFourComps
  The definitions reveal two interesting properties:
  \begin{enumerate}
  \item
    If we compose two maps \ACode{ϕ₁} and \ACode{ϕ₂}, then the
    resulting map is a renaming, iff both \ACode{ϕ₁} and \ACode{ϕ₂} are
    renamings.
    In other words:
    if \ACode{ϕ₁} is a \ACode{K₁}-map and \ACode{ϕ₂} is a \ACode{K₂}-map,
    then the result is a \ACode{(K₁ ⊔ K₂)}-map, where \ACode{⊔} refers to the
    lattice for \ACode{\{ \ADef{Kᵣ} , \ADef{Kₛ} \}}
    generated by \ACode{\ADef{Kᵣ} $\leq$ \ADef{Kₛ}}.
  \item
    All four compositions first apply \ACode{ϕ₁} to \ACode{x},
    and then apply \ACode{ϕ₂} to the result.
    If \ACode{ϕ₁} is a renaming, then this result is another variable, but
    if \ACode{ϕ₁} is a substitution, then this result is a term.
  \end{enumerate}
  With the \ACode{\ACon{Kit}} abstraction, we can easily abstract over
  \ACode{ϕ₂} being a renaming or a substitution:
  \ExampleCode\KExTwoComps

@@

  \LibCode\KComposeKit

  Similar as for the \ACode{\ACon{Kit}}, we derive further definitions
  from the the axioms of a \ACode{\ACon{ComposeKit}}:
  \begin{itemize}
  \item
    We define map composition as motivated by the previous subsection:
    \LibCode\KComposition
  \item
    We show that the \ACode{\AField{\_\&/⋯\_}} operator behaves like
    \ACode{\AField{\_\&\_}} when applied to variables:
    \LibCode\KComposeKitAp
    % \LibCode\KComposeKitApProof

    % The \ACode{\AField{`/id}} on the left side is from \ACode{K₁⊔K₂}, whereas
    % the \ACode{\AField{`/id}} on the right side is from \ACode{K₂}.
  \item
    We show that lifting \ACode{\AField{\_↑\_}} distributes over
    composition:
    \LibCode\KDistLiftCompose
    % \LibCode\KDistLiftComposeProof
  \end{itemize}
  To make it easier to explicitly specify a \ACode{\ACon{ComposeKit}},
  we define the notation \ACode{ϕ₁ \ADef{·[} C \ADef{]} ϕ₂}
  to stand for \ACode{ϕ₁ \ADef{·ₖ} ϕ₂} for some
  \ACode{\ACon{ComposeKit}} \ACode{C}.
  % \LibCode\KComposeKitNotation

  We then define the \ACode{\ACon{ComposeTraversal}} structure,
  which provides the \ACode{\ACon{⋯-fusion}} for the compositions
  described by arbitrary \ACode{\ACon{ComposeKit}}s:
  \LibCode\KComposeTraversal

  Similar as for the \ACode{\ACon{Traversal}}, we derive further
  definitions from the the axiom of a \ACode{\ACon{ComposeTraversal}}:
  \begin{itemize}
  \item 
    A map \ACode{ϕ} followed by a weakening is equivalent to a
    weakening followed by \ACode{ϕ} that has been lifted over the weakened
    variable:
    \LibCode\KCommLiftWeaken
    % \LibCode\KCommLiftWeakenProof
    By applying \ACode{\AField{⋯-fusion}} to this lemma, we get the same
    statement for multiple map applications:
    \LibCode\KCommLiftWeakenTraverse
    % \LibCode\KCommLiftWeakenTraverseProof
  \item 
    A weakening followed by a singleton substitution cancel each other out:
    \LibCode\KWeakenCancelsSingle
    % \LibCode\KWeakenCancelsSingleProof
    By applying \ACode{\AField{⋯-fusion}} and \ACode{\AField{⋯-id}} to
    this lemma, we get the same statement for multiple map applications:
    \LibCode\KWeakenCancelsSingleTraverse
    % \LibCode\KWeakenCancelsSingleTraverseProof
  \item 
    A singleton substitution followed by some map \ACode{ϕ} is the same
    as \ACode{ϕ} lifted over the variable of the singleton substitution
    followed by the singleton substitution with \ACode{ϕ} applied to its term:
    \LibCode\KDistLiftSingle
    % \LibCode\KDistLiftSingleProof
    By applying \ACode{\AField{⋯-fusion}} to this lemma, we get the
    same statement for multiple map applications:
    \LibCode\KWeakenCancelsSingleTraverse
    \LibCode\KDistLiftSingleTraverse
    % \LibCode\KDistLiftSingleTraverseProof
  \item 
    Given the \ACode{\ADef{⋯-↑-wk}} lemma, we can define the four
    \ACode{\ACon{ComposeKit}} instances:
    \LibCode\KComposeKitInstances
    \LibCode\KComposeKitInstancesConcrete
  \end{itemize}

  Finally, we instantiate the \ACode{\ACon{ComposeTraversal}} for
  System F to make the composition machinery available:
  \AppCode\FFusion
  \AppCode\FFusionProofInteresting
  % \AppCode\FFusionProofRest
  \AppCode\FComposeTraversal
